---
title: "Save All 21 Opening Truths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Save All 21 Opening Truths}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, message=FALSE, warning=FALSE}
library(backgammonr)
library(knitr)
```

This notebook saves one proxy-truth file for each of the 21 unordered opening rolls.

The function used here is `bg_truth_state()`.

- `state` is the board we start from.
- `roll` is the realized opening roll.
- `budget` is the proxy-truth simulation budget.
- `simulation_policy = "random"` means the continuation play inside the rollout is random legal play.
- `reward_model = "scalar_payoff"` means each rollout is reduced to one scalar payoff in `[0, 1]`.
- `posterior_model = "beta_pseudo"` means the package uses its scalar pseudo-Beta model when summarizing the move values.
- `save_path` is the permanent `.rds` file that gets written for that roll.

```{r settings}
truth_budget <- 100000L
truth_seed_base <- 1000L
truth_n_cores <- 12L
truth_parallel <- TRUE
overwrite_truth_files <- FALSE

opening_truth_dir <- file.path("artifacts", "opening_truths")
dir.create(opening_truth_dir, recursive = TRUE, showWarnings = FALSE)

opening_rolls <- data.frame(
  die1 = c(1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 5L, 5L, 6L),
  die2 = c(1L, 2L, 3L, 4L, 5L, 6L, 2L, 3L, 4L, 5L, 6L, 3L, 4L, 5L, 6L, 4L, 5L, 6L, 5L, 6L, 6L)
)

opening_rolls$roll_label <- paste0(opening_rolls$die1, "-", opening_rolls$die2)
opening_rolls$save_path <- file.path(
  opening_truth_dir,
  paste0("opening_", opening_rolls$roll_label, "_truth_budget_", truth_budget, ".rds")
)

opening_board <- bg_initial_board()
```

```{r opening-board}
bg_print_board(opening_board)
```

```{r all-opening-truths}
opening_truths <- vector("list", nrow(opening_rolls))
names(opening_truths) <- opening_rolls$roll_label

for (i in seq_len(nrow(opening_rolls))) {
  die1 <- opening_rolls$die1[[i]]
  die2 <- opening_rolls$die2[[i]]
  roll_label <- opening_rolls$roll_label[[i]]
  save_path <- opening_rolls$save_path[[i]]

  opening_truths[[i]] <- bg_truth_state(
    state = opening_board,
    roll = bg_roll(die1, die2),
    budget = truth_budget,
    simulation_policy = "random",
    reward_model = "scalar_payoff",
    posterior_model = "beta_pseudo",
    n_cores = truth_n_cores,
    parallel = truth_parallel,
    cache = FALSE,
    save_path = save_path,
    overwrite = overwrite_truth_files,
    seed = truth_seed_base + i - 1L,
    problem_id = paste0("opening_", roll_label)
  )
}
```

```{r saved-files}
knitr::kable(
  opening_rolls,
  caption = "Each opening roll is saved to its own permanent .rds file."
)
```

```{r opening-truth-summary}
opening_truth_summary <- do.call(
  rbind,
  lapply(opening_truths, function(x) x$summary)
)

opening_truth_summary <- merge(
  opening_truth_summary,
  opening_rolls[, c("roll_label", "die1", "die2", "save_path")],
  by.x = "roll",
  by.y = "roll_label",
  all.x = TRUE,
  sort = FALSE
)

opening_truth_summary <- opening_truth_summary[
  order(opening_truth_summary$die1, opening_truth_summary$die2),
  c(
    "problem_id",
    "roll",
    "die1",
    "die2",
    "reference_budget",
    "proxy_reference_best_move_label",
    "top_two_gap_estimate",
    "top_two_gap_mc_lower_95",
    "mc_gap_excludes_zero",
    "difficulty_label",
    "save_path"
  )
]

knitr::kable(
  opening_truth_summary,
  digits = 4,
  caption = "Summary of the 21 saved opening-roll proxy truths."
)
```

```{r opening-1-6-example}
truth_1_6_path <- file.path(
  opening_truth_dir,
  paste0("opening_1-6_truth_budget_", truth_budget, ".rds")
)

truth_1_6 <- bg_truth_load(truth_1_6_path)

knitr::kable(
  truth_1_6$summary[, c(
    "problem_id",
    "reference_budget",
    "proxy_reference_best_move_label",
    "top_two_gap_estimate",
    "top_two_gap_mc_lower_95",
    "mc_gap_excludes_zero",
    "difficulty_label"
  )],
  digits = 4,
  caption = "Saved proxy-truth summary for opening roll 1-6."
)
```

```{r opening-1-6-truth-plot, fig.width=8, fig.height=5}
plot_bg_truth(truth_1_6)
```

```{r opening-1-6-truth-table}
truth_1_6_table <- truth_1_6$reference$action_table[
  order(truth_1_6$reference$action_table$rank),
  c(
    "rank",
    "candidate_index",
    "move_label",
    "reference_mean",
    "reference_mc_lower_95",
    "reference_mc_upper_95",
    "allocation_count",
    "unresolved_fraction"
  )
]

knitr::kable(
  truth_1_6_table,
  digits = 4,
  caption = "Move table for the saved 1-6 opening-roll truth."
)
```



```{r}

truth_1_6 <- bg_truth_state(
  state = bg_initial_board(),
  roll = bg_roll(1L, 6L),
  budget = 100000L,
  simulation_policy = "random",
  n_cores = 12L,
  parallel = TRUE,
  cache = FALSE,
  save_path = file.path("artifacts", "opening_truths", "opening_1-6_truth_budget_100000.rds"),
  overwrite = FALSE,
  seed = 1005L,
  problem_id = "opening_1-6"
)



```




