---
title: "Thompson Sampling Theory for This Project"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thompson Sampling Theory for This Project}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# 1. The right theoretical frame

The package is not studying Thompson sampling in the most generic textbook
sense. It is studying Thompson sampling as a **finite-budget move-allocation
rule** on a fixed local backgammon decision problem.

Fix a board state $s$ and a realized roll $r$. Let the legal actions be

$$
A(s, r) = \{a_1, \dots, a_K\}.
$$

For each action $a_i$, define a rollout reward $Y_i \in [0, 1]$ and mean value

$$
\mu_i = \mathbb{E}[Y_i].
$$

The best action under the rollout model is

$$
a^* = \arg\max_{i \in \{1,\dots,K\}} \mu_i.
$$

A finite-budget method receives a total budget $N$ rollouts and must return a
recommendation $\hat{a}_N$.

So the question is not “how do we maximize cumulative reward forever?” The
question is closer to **best-arm identification** or **ranking and selection**:

- with limited simulation effort,
- and only one final recommendation,
- how should we allocate samples across the arms?

That distinction matters because it changes what “good behavior” looks like.
Cumulative-reward bandit logic and finite-budget recommendation logic are not
identical.

# 2. Why Thompson sampling fits this problem

Thompson sampling is attractive here because it turns posterior uncertainty into
an allocation rule without requiring a brittle deterministic threshold.

At step $t$:

1. maintain a posterior over each action value $\mu_i$,
2. sample a possible value from each posterior,
3. allocate the next rollout to the action whose sampled value is largest.

In words: sample according to the current posterior story of “which action might
really be best.”

That is a natural fit for move evaluation because finite-budget decision quality
depends heavily on whether the algorithm keeps revisiting actions that are still
plausibly optimal.

# 3. The package’s posterior model

The package uses a Beta-style working posterior for bounded rollout rewards.

Let the prior for action $i$ be

$$
\mu_i \sim \mathrm{Beta}(\alpha_0, \beta_0).
$$

After observing rollout outcomes $y_{i,1}, \dots, y_{i,n_i}$ with each
$y_{i,m} \in [0,1]$, the package updates using

$$
\alpha_i = \alpha_0 + \sum_{m=1}^{n_i} y_{i,m},
\qquad
\beta_i = \beta_0 + \sum_{m=1}^{n_i} (1 - y_{i,m}).
$$

The posterior mean is

$$
\mathbb{E}[\mu_i \mid \mathcal{D}] =
\frac{\alpha_i}{\alpha_i + \beta_i},
$$

and the Beta posterior variance is

$$
\mathrm{Var}(\mu_i \mid \mathcal{D}) =
\frac{\alpha_i \beta_i}
{(\alpha_i + \beta_i)^2 (\alpha_i + \beta_i + 1)}.
$$

## 3.1 Important nuance: this is a working posterior

This package is slightly more subtle than a pure Beta-Bernoulli story because
unresolved rollouts can contribute a fractional reward such as `0.5`. That
means the update remains mathematically Beta-shaped, but it is best interpreted
as a **working posterior for bounded rewards**, not as a literal exact conjugate
posterior for purely binary observations.

That is a good thing to explain explicitly. It avoids overstating the modeling
assumption.

The benefit of this working posterior is that it stays simple, fast, and
interpretable:

- posterior mean acts like a smoothed rollout-value estimate,
- posterior spread reflects remaining uncertainty,
- Thompson draws can still be sampled cheaply.

# 4. Thompson sampling as probability matching

If we could compute the posterior probability that action $i$ is optimal,

$$
p_i = \Pr\!\left(\mu_i = \max_j \mu_j \mid \mathcal{D}\right),
$$

then an intuitively appealing rule would be to sample the next action in
proportion to $p_i$.

Thompson sampling implements this idea indirectly. By drawing

$$
\tilde{\mu}_i \sim p(\mu_i \mid \mathcal{D})
$$

for every action and then choosing the action with the largest sampled value,
the long-run frequency with which an action is selected tracks its posterior
plausibility of being best.

That is why Thompson sampling is often described as a form of **probability
matching**.

For this package, that interpretation is more useful than generic
exploration-versus-exploitation language. The key idea is not merely “try
uncertain actions sometimes.” The key idea is:

> allocate rollout budget in proportion to how plausible it is that an action
> still matters for the final recommendation.

# 5. Why lower-mean actions can still deserve budget

Suppose action $a_1$ currently has the largest posterior mean. Why would
Thompson sampling ever allocate to some action $a_2$ with a lower mean?

Because the decision is not based only on mean. It is based on the **sampled**
draws from the posteriors. If $a_2$ is much more uncertain, then a sampled draw
from its posterior may still exceed the sampled draw for $a_1$.

That is exactly the kind of behavior we want in a finite-budget move evaluator:

- obviously bad actions should fade out,
- plausible challengers should keep receiving attention,
- uncertainty should not be ignored just because one action is temporarily ahead.

# 6. The algorithm in package terms

For one fixed board and roll, the package’s Thompson loop is conceptually:

1. generate legal actions,
2. initialize each action with a small number of rollouts,
3. compute Beta-style posterior parameters for each action,
4. repeat until the total budget is exhausted:
   - sample one posterior draw per action,
   - allocate the next rollout to the sampled leader,
   - update that action’s posterior,
5. rank actions by posterior mean and report the recommended move.

This logic appears in the allocation engine, but the important point for the
vignettes is that the package is doing something conceptually clean:

- **mechanics** create the candidate actions,
- **rollouts** generate noisy evidence,
- **posterior updates** summarize that evidence,
- **Thompson draws** decide where the next rollout goes.

# 7. What finite-budget behavior should we expect?

The project becomes much easier to interpret if we say ahead of time what
behaviors are normal.

## 7.1 Easy states with a clear top action

When the best action is well separated:

- posterior means separate quickly,
- weak actions lose posterior-optimality probability,
- Thompson allocation concentrates on the apparent leader,
- most methods eventually agree.

In these states Thompson may still save effort by avoiding unnecessary sampling
of obviously weak moves, but the qualitative recommendation will often be
stable.

## 7.2 Hard near-tie states

When two or more actions are close:

- posterior-optimality probabilities remain spread out,
- allocation remains more dispersed,
- finite-budget recommendations can flip,
- reference certification becomes much more important.

This is where Thompson sampling becomes most interesting. It should not look
perfectly stable too early. If the state is genuinely difficult, some instability
is the correct statistical behavior.

## 7.3 Large action sets

When there are many legal actions, a good allocation rule should discover that
most of them are weak and focus budget on the plausible leaders.

This is one reason Thompson sampling can be appealing in move-evaluation
problems. It does not need to commit to a hard deterministic elimination rule at
the start. Weak actions naturally fade as their posterior probability of being
best falls.

# 8. Early noise, overcommitment, and recovery

One of the most important finite-budget questions is whether a method can
recover from early luck.

A few lucky rollouts can make an action look strong at the beginning. If the
allocation rule becomes too deterministic too early, it may effectively lock in
to a bad action.

Thompson sampling is not immune to early noise, but it has an important
self-correcting feature: as long as an alternative action still has enough
posterior uncertainty to occasionally look best under posterior draws, it keeps
receiving at least some budget.

That does not guarantee perfect recovery. But it does create a more flexible
dynamic than a purely greedy mean-based rule.

This is exactly why the package includes trace functionality. A trace lets you
watch whether:

- one action takes over immediately,
- challengers continue to receive samples,
- the leader changes as evidence accumulates.

# 9. Top-two Thompson sampling (TTTS)

The package also includes top-two Thompson sampling.

The basic intuition is that in many best-action problems, the most useful
question is not “which action is sampled best among all actions?” but rather
“which challenger most deserves comparison with the current leader?”

TTTS emphasizes that leader-versus-challenger structure. In a hard near-tie
state, that can improve separation because it spends more time discriminating
among the top plausible actions.

The package exposes `evaluate_actions_ttts()` so that Thompson and TTTS can be
compared empirically rather than ideologically. That is the right framing:
TTTS is not automatically better, but it is often worth examining in hard
states.

# 10. Variance reduction and why it matters for Thompson studies

The package includes `dice_mode` options and common-random-number (`crn`)
control. These are not decorations. They matter because posterior updates depend
on rollout noise, and rollout noise can be shaped.

For example:

- stratified dice can reduce variance from the earliest continuation randomness,
- CRN can make action comparisons less noisy by aligning randomness across
  competing actions.

Those tools do not change the Thompson idea itself, but they can change how
quickly the posterior stabilizes. In other words, they affect the quality of the
evidence Thompson sampling receives.

# 11. How this theory maps to concrete package outputs

The most interpretable outputs in the package correspond directly to the
theoretical quantities above.

- `summary(th)` gives the one-row headline recommendation.
- `compare_action_posteriors(th)` gives a compact action table.
- `trace_thompson_allocation(...)` shows how allocation changed over time.
- `compare_thompson_to_reference(...)` compares a finite-budget run to a
  stronger proxy-truth run.
- `study_budget_tradeoff(...)` shows how regret and MSE change as the budget
  grows.

That means the theory does not live separately from the code. The code already
produces the objects that answer the theoretical questions.

# 12. A minimal theory-to-code example

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

th <- evaluate_actions_thompson(
  board = board,
  roll = roll,
  total_budget = 700L,
  rollout_policy = "random",
  max_rollout_turns = 250L,
  fast_diagnostics = FALSE,
  seed = 1L
)

summary(th)
compare_action_posteriors(th, top_n = 8L)
explain_move_evaluation(th)
```

Read this output as follows:

- the summary tells you what move was recommended,
- the posterior table tells you which other moves were still credible,
- the explanation turns the raw output into a sentence.

# 13. Why the problem is not arbitrary

A fair skeptical question is: why should anyone care about Thompson sampling on
this particular problem?

The answer is that the package isolates a very general situation:

- many alternatives,
- noisy finite-cost evaluation,
- one final decision,
- uncertainty that should guide sampling.

Backgammon is the environment, but the pattern appears everywhere. If a method
works well here, it suggests that the same logic may work in other local
decision problems with expensive simulation.

That does **not** mean the conclusions automatically transfer unchanged. It does
mean the project is studying a recognizable and important statistical structure.

# 14. How lessons here transfer elsewhere

The ideas in this vignette translate naturally to several other domains.

## 14.1 A/B testing with finite traffic

Different variants are analogous to legal moves. Traffic is analogous to
simulation budget. Thompson logic says: do not split traffic equally forever if
some variants are already implausible and others remain competitive.

## 14.2 Simulation optimization

Candidate system designs are analogous to legal moves. Each simulation run is
costly. Thompson sampling can direct effort toward designs that are still
plausibly best.

## 14.3 Hyperparameter tuning

Each hyperparameter configuration is a candidate action. Validation outcomes are
noisy. A finite evaluation budget must be allocated intelligently.

## 14.4 Recommender and ranking systems

A small set of leading candidates often deserves most of the evaluation effort.
Posterior-guided allocation is a principled way to focus attention while still
allowing challengers to compete.

## 14.5 Local search in larger sequential systems

Backgammon is especially suggestive here. Many larger search systems repeatedly
face local stochastic choices. The package is a controlled place to study what
an uncertainty-aware local allocator does.

# 15. What to look for in the workflow vignette

When you move to `05_thompson_workflow.Rmd`, the key questions to keep in mind
are the theoretical ones:

- Is the finite-budget recommendation stable?
- How concentrated was the allocation?
- Did Thompson focus on the plausible top actions?
- How close was the result to the high-budget reference?
- Does increasing the budget actually improve the decision?

Those questions keep the workflow tied to the main problem rather than turning
the code into a disconnected tour of function calls.

# 16. Self-contained near-tie instability diagnostic

Approx runtime:

- quick run (`budgets` length 4): about 15-60 seconds,
- deeper run (`budgets` length 6-8): about 1-3 minutes.

```{r}
library(backgammonr)

bg_plot_budget_stability(
  board = bg_initial_board(),
  roll = bg_roll(1L, 6L),
  budgets = c(64L, 128L, 256L, 512L, 1024L, 2048L),
  methods = c("thompson", "equal", "ucb"),
  rollout_policy = "random",
  max_rollout_turns = 250L,
  seed = 4L
)
```

Interpretation:

- recommendation flips at low budget are expected in hard states,
- what matters is whether instability decreases as budget increases.

# 17. Self-contained low-budget vs high-budget reference comparison

Approx runtime:

- quick run (`truth_budget = 4000`): about 30-120 seconds,
- deep run (`truth_budget = 30000`): about 4-12 minutes.

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

# Build one reference once, then reuse it.
ref <- approximate_action_reference(
  board = board,
  roll = roll,
  truth_budget = 12000L,     # deep run: 30000L
  rollout_policy = "random",
  max_rollout_turns = 300L,
  seed = 5L
)

cert <- certify_reference_truth(reference = ref)

cmp <- compare_thompson_to_reference(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = 300L,       # intentionally small-budget regime
  reference = ref,
  reference_certificate = cert,
  rollout_policy = "random",
  max_rollout_turns = 300L,
  seed = 6L
)

cmp$summary

cmp_clean <- cmp$action_table[, c(
  "move_label", "finite_recommended", "reference_best",
  "finite_allocation_count", "finite_estimate",
  "reference_estimate", "abs_error"
)]

head(cmp_clean[order(-cmp_clean$abs_error), ], 8L)
```

Interpretation:

- this isolates finite-budget Thompson against a stronger proxy reference,
- large `abs_error` rows identify actions where finite information is still weak,
- `proxy_pcs = 0` with tiny `simple_regret` can occur in near-tie states.

# 18. What to report in Thompson studies

At minimum, include:

1. budget and rollout-model settings,
2. recommendation plus uncertainty diagnostics,
3. allocation dynamics evidence,
4. finite-vs-reference PCS/regret/MSE metrics,
5. baseline comparisons,
6. runtime.

Without all six, conclusions about Thompson behavior are usually incomplete.

# 19. Assumptions and cautions

All results are conditional on:

- rollout policy,
- horizon truncation and unresolved-value treatment,
- posterior/prior specification,
- variance-control settings (`dice_mode`, CRN),
- reference-estimate quality.

Claims should be phrased conditionally and paired with sensitivity checks.
