---
title: "Thompson Workflow: From One Decision to Broader Evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thompson Workflow: From One Decision to Broader Evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette is the slow, Thompson-centered walk through the package.

The main idea is to stay tied to the actual research question:

> on one fixed local backgammon decision problem, what does finite-budget
> Thompson sampling do, how should we interpret it, and how does it compare with
> stronger reference evidence?

# 1. Set the problem up once

To avoid redundant code, we define the decision instance and the main simulation
settings only once.

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

rollout_policy <- "random"
max_rollout_turns <- 250L
seed <- 1L

total_budget <- 1200L
reference_budget <- 12000L   # deep run: use 30000L when feasible
budget_grid <- c(128L, 256L, 512L, 1024L, 2048L)
n_matchup_games <- 400L      # deep run: use 800L+ when feasible
```

This is the core pattern for the rest of the vignette:

- keep the board fixed,
- keep the roll fixed,
- keep the rollout model fixed,
- vary only the allocation budget or comparison method when needed.

Approx runtime guidance:

- quick/demo (`reference_budget = 4000`): roughly 1-4 minutes,
- default (`reference_budget = 12000`): roughly 3-10 minutes,
- deep (`reference_budget = 30000`): roughly 8-25 minutes.

# 2. Start with the legal action set

Before running Thompson sampling, inspect the action set it will allocate over.

```{r}
legal <- generate_legal_moves(board, roll)

legal_view <- summarize_legal_moves(
  legal,
  max_candidates = 8L,
  print_table = FALSE
)

legal_view
```

This matters because Thompson sampling is not choosing from a vague search
space. It is choosing among these concrete legal actions.

# 3. Run one finite-budget Thompson evaluation

```{r}
th <- evaluate_actions_thompson(
  board = board,
  roll = roll,
  total_budget = total_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed
)

summary(th)
```

The one-row summary answers the first practical question: what move was
recommended at this budget?

# 4. Look at the top posterior summaries

A one-row summary is not enough on its own. We also want to know which actions
were still competitive.

```{r}
posterior_view <- compare_action_posteriors(th, top_n = 8L)

posterior_view[, c(
  "move_label",
  "recommended",
  "allocation_count",
  "estimate",
  "uncertainty_sd",
  "prob_best",
  "exp_regret"
)]
```

This is the main compact table to read.

Interpret the columns this way:

- `move_label`: human-readable action,
- `recommended`: whether this is the current leader,
- `allocation_count`: how much budget Thompson gave it,
- `estimate`: posterior mean estimate,
- `uncertainty_sd`: remaining uncertainty,
- `prob_best`: how plausible it still is that this action is best,
- `exp_regret`: posterior expected regret if chosen.

# 5. Turn the output into a sentence

```{r}
explain_move_evaluation(th)
```

This is a small but useful bridge between the statistical object and the human
reading experience. Good workflow documents should use functions like this more,
not less.

# 6. Watch the allocation evolve

A big advantage of this package is that it can expose the dynamics, not just the
final answer.

```{r}
trace <- trace_thompson_allocation(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = total_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  trace_every = 40L,
  seed = seed
)

trace$checkpoint_summary[, c(
  "checkpoint",
  "selected_move_label",
  "leader_move_label",
  "leader_estimate",
  "leader_posterior_sd",
  "leader_allocation_count"
)]
```

This table is the easiest place to see whether Thompson sampling is behaving the
way the theory vignette suggested:

- does one action dominate early?
- do challengers keep getting budget?
- does the leader stabilize?

# 7. Build a stronger proxy reference

A finite-budget Thompson run is only part of the story. We also want a stronger
reference estimate so that we can talk about regret and approximation quality.

```{r}
reference <- approximate_action_reference(
  board = board,
  roll = roll,
  truth_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed
)

certificate <- certify_reference_truth(reference = reference)

certificate$certificate
```

This is where the workflow stays honest. The reference is stronger than the
finite-budget run, but it is still simulated, so we check how well separated the
top reference actions appear to be.

# 8. Compare the finite-budget Thompson run with the reference

```{r}
comparison <- compare_thompson_to_reference(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = 512L,
  reference = reference,
  reference_certificate = certificate,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 1L
)

comparison$summary
```

This one-row output is one of the most important in the package because it ties
the main problem together:

- finite budget,
- Thompson recommendation,
- stronger proxy reference,
- regret and error metrics,
- reference difficulty.

# 9. Inspect the action-level comparison, but keep it compact

```{r}
action_view <- comparison$action_table[order(
  -comparison$action_table$finite_allocation_count
), ]

action_view[, c(
  "move_label",
  "finite_recommended",
  "reference_best",
  "finite_allocation_count",
  "finite_estimate",
  "reference_estimate",
  "abs_error"
)][1:8, ]
```

This is much easier to read than dumping every column. It directly answers:

- what did Thompson focus on?
- did the recommended move match the reference-best move?
- where was the remaining error concentrated?

# 10. Study the budget tradeoff on the same problem

One of the most important questions in the project is whether more budget really
improves recommendation quality in a meaningful way.

```{r}
tradeoff <- study_budget_tradeoff(
  board = board,
  roll = roll,
  method = "thompson",
  budgets = budget_grid,
  truth_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed + 2L
)

tradeoff$results[, c(
  "total_budget",
  "chosen_move_label",
  "truth_best_move_label",
  "correct_selection",
  "simple_regret",
  "mse",
  "runtime_seconds"
)]
```

This is the clearest budget story in the package. It lets you see:

- whether the chosen move stabilizes,
- whether simple regret shrinks,
- whether MSE shrinks,
- what runtime you paid for those improvements.

# 11. Compare Thompson with baseline methods on the same budget

The package should not present Thompson in isolation. Baselines matter.

```{r}
methods <- compare_methods_on_position(
  board = board,
  roll = roll,
  methods = c("thompson", "equal", "ucb", "ocba", "greedy"),
  total_budget = total_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed + 3L
)

methods$summary[, c(
  "method",
  "recommended_move_label",
  "recommended_estimate",
  "recommended_prob_best",
  "recommended_expected_regret",
  "recommended_allocation_count",
  "runtime_seconds"
)]
```

This does not settle the whole research question, but it is a very useful same-
position, same-budget snapshot.

# 12. Use Thompson as a player in many full games

The single-decision workflow is the scientific center, but it is also useful to
see what happens when Thompson-based move choice is embedded in many full games.

```{r}
thompson_matchup <- simulate_matchup(
  player1 = "thompson_rollout",
  player2 = "random",
  n_games = n_matchup_games,
  board = board,
  max_turns = 150L,
  rollout_budget = 16L,
  rollout_policy = rollout_policy,
  max_rollout_turns = 150L,
  seed = seed + 4L
)

summary(thompson_matchup)[, c(
  "n_games",
  "completed_games",
  "player1_wins",
  "player2_wins",
  "player1_win_rate",
  "player2_win_rate",
  "mean_turns"
)]
```

This is not the same question as the single-position evaluation. It is a broader
application question: if Thompson-style move selection is used repeatedly, what
kind of game-level behavior emerges?

# 13. Variance-control sensitivity on the same instance

Noise-control settings can change how quickly posterior quantities stabilize.
This section keeps the same board and roll but varies `dice_mode` and `crn`.

```{r}
var_study <- study_variance_controls(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = 1200L,
  dice_modes = c("iid", "stratified_first_roll", "stratified_first_two_rolls"),
  crn_values = c(FALSE, TRUE),
  truth_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 5L
)

var_study$results[, c(
  "dice_mode",
  "crn",
  "chosen_move_label",
  "correct_selection",
  "simple_regret",
  "mse",
  "runtime_seconds"
)]
```

Interpret this as a sensitivity check, not a one-off winner table.

# 14. Multi-case benchmark with enough simulation calls

One instance is useful for interpretation, but robust behavior claims require
multiple cases and larger total simulation counts.

```{r}
cases <- list(
  bg_benchmark_case(bg_initial_board(), bg_roll(1L, 6L), case_id = "init_1_6"),
  bg_benchmark_case(bg_initial_board(), bg_roll(3L, 2L), case_id = "init_3_2"),
  bg_benchmark_case(bg_initial_board(), bg_roll(5L, 4L), case_id = "init_5_4"),
  bg_benchmark_case(bg_initial_board(), bg_roll(2L, 1L), case_id = "init_2_1"),
  bg_benchmark_case(bg_initial_board(), bg_roll(6L, 6L), case_id = "init_6_6"),
  bg_benchmark_case(bg_initial_board(), bg_roll(4L, 3L), case_id = "init_4_3")
)

bm <- benchmark_thompson(
  cases = cases,
  budgets = c(256L, 1024L, 4096L),
  baselines = c("equal", "ucb", "ocba", "greedy"),
  include_ttts = TRUE,
  reference_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 6L
)

print(bm)

focus <- summarize_thompson_benchmark(bm)
head(focus$thompson, 12L)
head(focus$relative_to_thompson, 20L)
head(focus$by_difficulty, 20L)
```

This loop is intentionally substantial: multiple cases, multiple budgets, and
multiple baselines so state-dependent behavior is visible.

# 15. Function-coverage map

The workflow above covers the main user-facing path:

- mechanics: `bg_initial_board()`, `bg_roll()`, `generate_legal_moves()`,
  `summarize_legal_moves()`, `apply_move()`
- finite-budget evaluation: `evaluate_actions_thompson()`,
  `compare_action_posteriors()`, `summary()`, `explain_move_evaluation()`
- dynamics: `trace_thompson_allocation()`, `plot_thompson_convergence()`
- reference layer: `approximate_action_reference()`,
  `certify_reference_truth()`, `compare_thompson_to_reference()`
- budget studies: `study_budget_tradeoff()`, `study_variance_controls()`
- benchmarks: `compare_methods_on_position()`, `benchmark_thompson()`,
  `summarize_thompson_benchmark()`
- game-level application: `simulate_matchup()`

# 16. Result-discussion outline

For reports and presentations, cover results in this order:

1. define instance and action-set context,
2. state budget and rollout-model settings,
3. report recommendation with uncertainty,
4. report finite-vs-reference metrics (PCS/regret/MSE),
5. show allocation-dynamics evidence,
6. compare against baselines at matched budgets,
7. report runtime tradeoffs,
8. state limitations (reference uncertainty, difficulty, model dependence).

# 17. Future optimization and model directions

Useful next directions (hypotheses, not claims):

- multi-core rollout batching for large benchmark grids,
- richer posterior models for bounded non-Bernoulli rewards,
- expanded TTTS/gap-aware/stopping-rule variants,
- CRN-aware Bayesian allocation refinements,
- tighter reference-certification diagnostics.

# 18. What this workflow is trying to teach

The workflow is intentionally ordered.

1. fix the decision instance,
2. inspect the legal actions,
3. run finite-budget Thompson,
4. inspect the posterior table,
5. trace the allocation,
6. build a stronger reference,
7. compare finite budget against reference,
8. study the budget tradeoff,
9. compare with baselines,
10. embed the method in many games.

That order matters because it keeps every code block tied to the main problem.
The goal is not to show more functions than necessary. The goal is to make the
research question legible.

# 19. A final reading rule

Whenever a Thompson result looks impressive, ask:

- was the reference credible?
- was the state easy or hard?
- did the result persist as the budget increased?
- did the method still look strong against the baselines?

Whenever a Thompson result looks unstable, ask:

- was the state a genuine near tie?
- did the trace show early luck and later correction?
- would a larger budget or stronger variance control change the picture?

That is the mindset this package is built to support.
