---
title: "Main Question Example: Thompson at Finite Budget vs High Reference"
output: rmarkdown::html_vignette
params:
  run_code: false
  quick_mode: true
vignette: >
  %\VignetteIndexEntry{Main Question Example: Thompson at Finite Budget vs High Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
if (!exists("params")) {
  params <- list(run_code = FALSE, quick_mode = TRUE)
}
run_code <- isTRUE(params$run_code)
quick_mode <- isTRUE(params$quick_mode)
options(bg_quick_mode = quick_mode)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_code,
  fig.width = 7.5,
  fig.height = 4.8
)
```

This document is a single, self-contained research workflow focused on one
question:

> with a finite rollout budget, does Thompson sampling allocate simulation
> effort in a statistically sensible way, and how close does it get to a
> stronger high-budget reference estimate?

It is centered on these functions:

- `bg_initial_board()`
- `bg_roll()`
- `generate_legal_moves()`
- `evaluate_actions_thompson()`
- `approximate_action_reference()`
- `compare_thompson_to_reference()`
- `study_budget_tradeoff()`
- `study_variance_controls()`
- `summarize_thompson_benchmark()`

Additional helper functions are included where they improve interpretation:

- `bg_print_board()`, `bg_plot_board()`
- `compare_action_posteriors()`, `explain_move_evaluation()`
- `trace_thompson_allocation()`, `plot_thompson_convergence()`
- `compare_methods_on_position()`
- `bg_plot_budget_stability()`
- `benchmark_thompson()`, `plot_thompson_vs_baselines()`

# 1. Statistical framing before code

Fix one local decision instance $d = (s, r)$ with board state $s$ and realized
roll $r$. Let legal actions be $A_d = \{1, \dots, K\}$. For each action $i$,
rollout reward $Y_i \in [0,1]$ has mean

$$
\mu_i = \mathbb{E}[Y_i].
$$

The rollout-model best action is

$$
a^* = \arg\max_i \mu_i.
$$

A finite-budget method uses total budget $N$ and returns recommendation
$\hat a_N$. Core evaluation quantities are:

- probability of correct selection (PCS):
$$
\Pr(\hat a_N = a^*)
$$
- simple regret:
$$
\mu_{a^*} - \mu_{\hat a_N}
$$
- value-estimation error, often summarized by MSE.

In this package, we approximate $a^*$ using a high-budget reference estimate,
not exact truth.

# 2. Thompson intuition in this package

Thompson sampling maintains posterior uncertainty for each action value, samples
one draw per action, and allocates the next rollout to the sampled leader.

Conceptually:

1. maintain posterior for each $\mu_i$,
2. draw $\tilde\mu_i$ from each posterior,
3. allocate next rollout to $\arg\max_i \tilde\mu_i$,
4. update posterior for sampled action.

Why this matters here:

- clearly weak actions should lose allocation over time,
- plausible leaders should keep receiving budget,
- hard near-tie states should show slower stabilization.

# 3. Set one fixed decision instance and budgets

Approx runtime guidance (laptop-class machine):

- quick run: `reference_budget = 12000` and smaller budget grids,
- deep run: `reference_budget = 30000` and larger budget grids.

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

rollout_policy <- "random"
max_rollout_turns <- if (isTRUE(getOption("bg_quick_mode", TRUE))) 250L else 300L
seed <- 11L

finite_budget <- if (isTRUE(getOption("bg_quick_mode", TRUE))) 1200L else 2000L
reference_budget <- if (isTRUE(getOption("bg_quick_mode", TRUE))) 12000L else 30000L

budget_grid <- if (isTRUE(getOption("bg_quick_mode", TRUE))) {
  c(128L, 256L, 512L, 1024L, 2048L)
} else {
  c(128L, 256L, 512L, 1024L, 2048L, 4096L)
}

benchmark_budgets <- if (isTRUE(getOption("bg_quick_mode", TRUE))) c(256L, 1024L) else c(256L, 1024L, 4096L)
trace_every <- if (isTRUE(getOption("bg_quick_mode", TRUE))) 120L else 80L
```

Interpretation: all comparisons below are conditioned on the same `board`,
`roll`, rollout policy, and horizon.

# 4. Print and plot the board and roll

```{r}
bg_print_board(board)
print(roll)
bg_plot_board(board)
```

Use both:

- `bg_print_board()` for exact textual state,
- `bg_plot_board()` for quick visual orientation.

# 5. Generate legal actions and inspect action space

```{r}
legal <- generate_legal_moves(board, roll)

legal_view <- summarize_legal_moves(
  legal,
  max_candidates = 10L,
  print_table = FALSE
)

legal_view[, c("candidate_index", "move_label", "n_steps")]

# Optional visual move diagnostics
bg_plot_legal_moves(board, roll)
```

Interpretation question: how large and how structured is the action set that
Thompson must allocate over?

# 6. Run one finite-budget Thompson evaluation

```{r}
th <- evaluate_actions_thompson(
  board = board,
  roll = roll,
  total_budget = finite_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed
)

print(th)
summary(th)
compare_action_posteriors(th, top_n = 8L)[, c(
  "move_label", "recommended", "allocation_count",
  "estimate", "uncertainty_sd", "prob_best", "exp_regret"
)]

explain_move_evaluation(th)
```

Interpretation:

- `allocation_count`: where finite budget went,
- `estimate` + `uncertainty_sd`: current value and uncertainty,
- `prob_best`: posterior plausibility that action is best,
- `exp_regret`: posterior downside if that action were chosen.

# 7. Compare Thompson against baseline methods at same budget

This gives a same-instance, same-budget cross-method snapshot.

```{r}
methods <- compare_methods_on_position(
  board = board,
  roll = roll,
  methods = c("thompson", "equal", "ucb", "ocba", "greedy", "ttts"),
  total_budget = finite_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed + 1L
)

methods$summary[, c(
  "method",
  "recommended_move_label",
  "recommended_estimate",
  "recommended_prob_best",
  "recommended_expected_regret",
  "recommended_allocation_count",
  "estimate_gap_top2",
  "runtime_seconds"
)]
```

Interpretation: this is not a final benchmark, but it is a clean first check of
whether Thompson is competitive on this specific state.

# 8. Trace Thompson allocation dynamics over time

```{r}
tr <- trace_thompson_allocation(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = finite_budget,
  trace_every = trace_every,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 2L
)

tr$checkpoint_summary[, c(
  "checkpoint",
  "selected_move_label",
  "leader_move_label",
  "leader_estimate",
  "leader_posterior_sd",
  "leader_allocation_count"
)]

plot_thompson_convergence(tr, metric = "allocation_count")
plot_thompson_convergence(tr, metric = "estimate")
```

Interpretation questions:

- does allocation concentrate too early?
- do challengers keep some budget in hard states?
- does the leader stabilize as evidence accumulates?

# 9. Plot recommendation stability across budgets

This is an explicit instability/mistake diagnostic.

```{r}
bg_plot_budget_stability(
  board = board,
  roll = roll,
  budgets = budget_grid,
  methods = c("thompson", "equal", "ucb"),
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 3L
)
```

Interpretation:

- low-budget flips are expected in hard states,
- what matters is whether instability shrinks as budget grows.

# 10. Build a high-budget reference estimate and certify it

```{r}
ref <- approximate_action_reference(
  board = board,
  roll = roll,
  truth_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed
)

cert <- certify_reference_truth(reference = ref)
cert$certificate
```

Important interpretation rule: reference is a proxy truth. Certification helps
flag when top-two separation is weak.

# 11. Compare finite Thompson output to the reference

```{r}
cmp <- compare_thompson_to_reference(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = finite_budget,
  reference = ref,
  reference_certificate = cert,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 4L
)

cmp$summary[, c(
  "method", "total_budget", "reference_budget",
  "proxy_pcs", "simple_regret", "mse",
  "finite_runtime_seconds", "reference_runtime_seconds",
  "difficulty_label", "reference_certified"
)]

cmp_actions <- cmp$action_table[order(-cmp$action_table$finite_allocation_count), c(
  "move_label", "finite_recommended", "reference_best",
  "finite_allocation_count", "finite_estimate",
  "reference_estimate", "abs_error"
)]

head(cmp_actions, 10L)

plot(
  cmp_actions$finite_allocation_count,
  cmp_actions$abs_error,
  pch = 19,
  col = ifelse(cmp_actions$finite_recommended, "#2E86AB", "#444444"),
  xlab = "Finite allocation count",
  ylab = "Absolute error vs reference",
  main = "Where finite-budget error remains"
)
grid()

# Richer finite-vs-reference diagnostics
cmp_diag <- cmp$action_table[, c(
  "move_label", "finite_recommended", "reference_best",
  "finite_allocation_count", "finite_estimate", "finite_posterior_sd",
  "reference_estimate", "abs_error"
)]

old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# (1) Calibration-style scatter: finite estimate vs reference estimate
plot(
  cmp_diag$reference_estimate,
  cmp_diag$finite_estimate,
  pch = 19,
  col = ifelse(cmp_diag$finite_recommended, "#2E86AB", "#444444"),
  xlab = "Reference estimate",
  ylab = "Finite Thompson estimate",
  main = "Finite vs reference (action-level)"
)
abline(0, 1, lty = 2, col = "red")
text(
  cmp_diag$reference_estimate,
  cmp_diag$finite_estimate,
  labels = seq_len(nrow(cmp_diag)),
  pos = 3,
  cex = 0.7
)

# (2) Sorted absolute errors by action
ord_err <- order(cmp_diag$abs_error, decreasing = TRUE)
err_sorted <- cmp_diag[ord_err, , drop = FALSE]
plot(
  err_sorted$abs_error,
  seq_len(nrow(err_sorted)),
  pch = 19,
  yaxt = "n",
  xlab = "Absolute error",
  ylab = "Action",
  main = "Largest finite-vs-reference errors"
)
segments(0, seq_len(nrow(err_sorted)), err_sorted$abs_error, seq_len(nrow(err_sorted)))
axis(2, at = seq_len(nrow(err_sorted)), labels = err_sorted$move_label, las = 1, cex.axis = 0.65)

# (3) Posterior uncertainty vs absolute error
plot(
  cmp_diag$finite_posterior_sd,
  cmp_diag$abs_error,
  pch = 19,
  cex = pmax(0.7, sqrt(cmp_diag$finite_allocation_count / max(cmp_diag$finite_allocation_count))),
  col = ifelse(cmp_diag$reference_best, "#C0392B", "#444444"),
  xlab = "Finite posterior SD",
  ylab = "Absolute error vs reference",
  main = "Uncertainty-error relationship"
)
legend(
  "topright",
  legend = c("Reference-best action", "Other actions"),
  col = c("#C0392B", "#444444"),
  pch = 19,
  bty = "n",
  cex = 0.8
)

# (4) Rank agreement: finite ranking vs reference ranking
rank_df <- cmp_diag
rank_df$finite_rank <- rank(-rank_df$finite_estimate, ties.method = "first")
rank_df$reference_rank <- rank(-rank_df$reference_estimate, ties.method = "first")
plot(
  rank_df$reference_rank,
  rank_df$finite_rank,
  pch = 19,
  col = ifelse(rank_df$finite_recommended, "#2E86AB", "#444444"),
  xlab = "Reference rank (1 = best)",
  ylab = "Finite rank (1 = best)",
  main = "Rank agreement"
)
abline(0, 1, lty = 2, col = "red")
text(rank_df$reference_rank, rank_df$finite_rank, labels = seq_len(nrow(rank_df)), pos = 3, cex = 0.7)

par(old_par)
```

Interpretation:

- `proxy_pcs`: did finite recommendation match reference-best?
- `simple_regret`: decision loss relative to reference-best value,
- `mse`: whole action-vector estimation error,
- error-vs-allocation plot: where finite information is still weak,
- calibration scatter: whether finite action estimates track reference values,
- sorted-error chart: which specific actions are hardest at this budget,
- uncertainty-error plot: whether high posterior uncertainty aligns with larger
  reference gaps,
- rank-agreement plot: whether finite ordering is close to reference ordering.

# 12. Study budget tradeoff directly

```{r}
trade <- study_budget_tradeoff(
  board = board,
  roll = roll,
  method = "thompson",
  budgets = budget_grid,
  truth_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed + 5L
)

trade_view <- trade$results[, c(
  "total_budget",
  "chosen_move_label",
  "truth_best_move_label",
  "correct_selection",
  "simple_regret",
  "mse",
  "runtime_seconds"
)]

trade_view

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

plot(
  trade_view$total_budget,
  as.numeric(trade_view$correct_selection),
  type = "b", pch = 19, ylim = c(-0.05, 1.05),
  xlab = "Budget", ylab = "Proxy PCS (0/1)", main = "Budget vs PCS"
)
plot(
  trade_view$total_budget,
  trade_view$simple_regret,
  type = "b", pch = 19,
  xlab = "Budget", ylab = "Simple regret", main = "Budget vs Regret"
)
plot(
  trade_view$total_budget,
  trade_view$mse,
  type = "b", pch = 19,
  xlab = "Budget", ylab = "MSE", main = "Budget vs MSE"
)
par(old_par)

if (all(c("chosen_estimate", "chosen_truth_value", "truth_best_value") %in% names(trade$results))) {
  gap_df <- trade$results[order(trade$results$total_budget), c(
    "total_budget", "chosen_estimate", "chosen_truth_value", "truth_best_value"
  )]
  gap_df$finite_ref_gap <- gap_df$chosen_estimate - gap_df$chosen_truth_value
  gap_df$to_ref_best_gap <- gap_df$truth_best_value - gap_df$chosen_truth_value

  old_par <- par(no.readonly = TRUE)
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  plot(
    gap_df$total_budget,
    gap_df$finite_ref_gap,
    type = "b", pch = 19,
    xlab = "Budget",
    ylab = "Chosen estimate - reference value",
    main = "Finite-reference value gap"
  )
  abline(h = 0, lty = 2, col = "red")
  plot(
    gap_df$total_budget,
    gap_df$to_ref_best_gap,
    type = "b", pch = 19,
    xlab = "Budget",
    ylab = "Reference regret to best action",
    main = "Gap to reference-best value"
  )
  par(old_par)
}
```

Interpretation:

- this is the key convergence story for one instance,
- finite-budget non-monotonicity can happen, focus on trend,
- include runtime when discussing quality gains,
- the gap plots show whether chosen finite values and chosen actions approach the
  reference trajectory as budget grows.

# 13. Check variance-control sensitivity

```{r}
var_study <- study_variance_controls(
  board = board,
  roll = roll,
  method = "thompson",
  total_budget = finite_budget,
  dice_modes = c("iid", "stratified_first_roll", "stratified_first_two_rolls"),
  crn_values = c(FALSE, TRUE),
  truth_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  fast_diagnostics = FALSE,
  seed = seed + 6L
)

var_study$results[, c(
  "dice_mode", "crn", "chosen_move_label",
  "correct_selection", "simple_regret", "mse", "runtime_seconds"
)]
```

Interpretation: if conclusions change radically across settings, report that
explicitly.

# 14. Run multi-case Thompson benchmark and summarize

To use `summarize_thompson_benchmark()`, first run `benchmark_thompson()`.

```{r}
cases <- list(
  bg_benchmark_case(bg_initial_board(), bg_roll(1L, 6L), case_id = "init_1_6"),
  bg_benchmark_case(bg_initial_board(), bg_roll(3L, 2L), case_id = "init_3_2"),
  bg_benchmark_case(bg_initial_board(), bg_roll(5L, 4L), case_id = "init_5_4"),
  bg_benchmark_case(bg_initial_board(), bg_roll(2L, 1L), case_id = "init_2_1"),
  bg_benchmark_case(bg_initial_board(), bg_roll(6L, 6L), case_id = "init_6_6"),
  bg_benchmark_case(bg_initial_board(), bg_roll(4L, 3L), case_id = "init_4_3")
)

if (isTRUE(getOption("bg_quick_mode", TRUE))) {
  cases <- cases[1:3]
}

bm <- benchmark_thompson(
  cases = cases,
  budgets = benchmark_budgets,
  baselines = c("equal", "ucb", "ocba", "greedy"),
  include_ttts = TRUE,
  reference_budget = reference_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed + 7L
)

focus <- summarize_thompson_benchmark(bm)

head(focus$thompson, 12L)
head(focus$relative_to_thompson, 20L)
head(focus$by_difficulty, 20L)

plot_thompson_vs_baselines(bm, metric = "correct_selection_rate")
plot_thompson_vs_baselines(bm, metric = "mean_simple_regret")
plot_thompson_vs_baselines(bm, metric = "mean_runtime_seconds")
```

Interpretation:

- this is where state-dependent strengths and weaknesses become visible,
- report both favorable and unfavorable cells,
- discuss quality and runtime together.

# 15. Interpretation checklist (what to say in results)

For each figure/table, answer explicitly:

1. what statistical question this output is answering,
2. whether the state appears easy/hard (`difficulty_label`, certification),
3. whether finite Thompson is converging toward reference behavior,
4. where errors remain and why they may remain,
5. whether gains persist against baselines,
6. what runtime cost was paid.

# 16. What counts as encouraging vs cautionary

Encouraging patterns:

- allocation increasingly concentrates on plausible best actions,
- regret and MSE generally decline with budget,
- proxy PCS improves and recommendation flips decrease,
- Thompson is competitive or stronger than baselines on many cells.

Cautionary patterns:

- persistent instability at moderate budgets,
- large finite-vs-reference gaps despite high finite budget,
- improvements that disappear outside one easy case,
- strong dependence on one variance-control setting,
- reference certification indicates unresolved near-tie ambiguity.

# 17. Final takeaway

This workflow is not about generating large tables. It is about
finite-budget statistical decision evaluation:

- define one concrete decision instance,
- study Thompson allocation behavior,
- compare to a stronger reference estimate,
- check robustness across budgets and variance controls,
- validate broader behavior across multiple benchmark cases.

That is the core research use of this package.
