---
title: "Project Motivation: Why This Package Exists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Project Motivation: Why This Package Exists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# 1. Start with the actual problem

This package is easiest to understand if we start from the statistical question,
not from the function list.

The central object is a **local decision problem** in stochastic backgammon:

- there is a current board state $s$,
- a roll $r$ has already been realized,
- the player faces a finite set of legal actions $A(s, r)$,
- each legal action can be evaluated by stochastic rollout.

For one legal action $a_i$, define its rollout value under a chosen continuation
policy as

$$
\mu_i = \mathbb{E}[Y_i],
$$

where $Y_i \in [0, 1]$ is the rollout reward. In this package, rollout rewards
are typically interpreted as a bounded win-value proxy:

- `1` for a rollout win,
- `0` for a rollout loss,
- `unresolved_value` (often `0.5`) if the rollout horizon is truncated before
  resolution.

So the scientific question is:

> with a finite simulation budget $N$, how should we distribute rollouts across
> legal actions in order to recommend a strong move?

That is a ranking-and-selection problem. It is also a Bayesian allocation
problem. It is also a Monte Carlo design problem. Backgammon is the laboratory
used to make that problem concrete and difficult in a realistic way.

# 2. Why the package is not “just about backgammon”

A reader can easily misread the package as “an R backgammon engine with some
sampling code on top.” That interpretation is too narrow.

The package is really about **finite-budget decision quality under noisy
simulation**. Backgammon matters because it creates exactly the kinds of local
decision problems that make adaptive allocation interesting:

- realized randomness from dice changes the action set from turn to turn,
- legal move sets vary in size and difficulty,
- continuation outcomes are noisy,
- rollout depth matters,
- many states are easy, but some are genuinely close.

This is a much better testbed than a toy multi-armed bandit with synthetic
Gaussian arms because the arms here are induced by real move sequences in a
stochastic sequential environment.

That matters for interpretation. The package is not trying to prove that
Thompson sampling is universally best in every abstract setting. It is asking a
more grounded question:

> when the decision problem is generated by actual game mechanics and
> continuation randomness, what does Thompson-style allocation do with a finite
> budget?

# 3. Why backgammon is a good testbed and not an arbitrary choice

Backgammon has several features that make it especially useful for this kind of
work.

## 3.1 Realized randomness enters before the decision

In many local game-analysis settings, a decision is conditioned on a realized
random event. In backgammon, that event is the roll. Once the roll is fixed,
the decision problem is

$$
d = (s, r),
$$

not just “what is the best move from board $s$ in the abstract?”

That means the package studies a clean, reproducible local choice. If we keep
the board and the roll fixed, all allocation methods are solving the **same**
problem.

## 3.2 The action set is structured, not artificial

The legal actions are generated by the game rules. They are not synthetic arms
invented for convenience. Some actions differ sharply; some differ only subtly.
Some positions produce a handful of candidates; others produce a much larger
set. That variation is useful because finite-budget methods often behave
differently when:

- the number of plausible actions is small,
- the best action is clearly separated,
- several near-tied actions compete for attention.

## 3.3 Rollout values are noisy in a realistic way

A rollout is not an abstract reward draw detached from context. It is the result
of continuing the game under some baseline policy. So the rollout noise is tied
to the underlying environment:

- dice introduce chance variation,
- continuation policy introduces model variation,
- finite rollout horizon introduces truncation effects.

This creates exactly the practical problem that adaptive allocation is supposed
to handle: we need to distinguish truly strong actions from merely lucky ones.

## 3.4 The package can support both one-instance and many-instance reasoning

A single board-plus-roll decision problem is useful for explanation. A set of
cases is useful for benchmarking. Repeated full games are useful for seeing how
a local decision rule behaves when embedded in a player.

That gives the package three levels of evidence:

1. **one decision instance**: how does the method allocate budget?
2. **reference comparison**: how close is the finite-budget recommendation to a
   high-budget proxy?
3. **many full games**: how does the policy behave when used repeatedly?

# 4. Why finite budgets matter so much

It would be easy to say “just run enough simulations and the best move will
eventually reveal itself.” That is true in a vague asymptotic sense and not
very useful in practice.

Real work happens under finite compute budgets. If we have only $N$ rollouts in
total, then we must choose an allocation vector

$$
(n_1, \dots, n_K)
\quad \text{such that} \quad
\sum_{i=1}^K n_i = N,
$$

where $K$ is the number of legal actions.

That is the whole problem.

Equal allocation is simple and often reasonable. But equal allocation can be
wasteful when some actions are already obviously weak and others are close. A
good adaptive rule should learn to spend more effort where uncertainty still
matters.

So the package is about the finite-budget regime where design choices are still
visible:

- small budgets can lead to unstable recommendations,
- moderate budgets reveal whether an adaptive rule learns quickly,
- large budgets show whether finite-budget gains persist or disappear.

# 5. What “best move” means in this package

A major source of confusion in projects like this is that the phrase “best move”
sounds absolute. In this package it is not absolute.

The package studies the best move **under the current rollout model**:

- with a chosen rollout policy,
- with a chosen maximum rollout horizon,
- with a chosen unresolved-value convention.

So when we estimate $\mu_i$, we are not claiming exact game-theoretic truth.
We are estimating the action value induced by the package’s rollout design.

That is still valuable. It lets us ask whether finite-budget Thompson sampling
tracks the best action under a fixed evaluation model. But it also means the
project should be described honestly:

- this is not a solved-backgammon oracle,
- this is not exact equilibrium analysis,
- this is a finite-budget simulation-allocation framework.

That honesty makes the problem less arbitrary, not more. Once the object of
study is clearly defined, the conclusions become more interpretable.

# 6. Why proxy truth is necessary

The exact action values $\mu_i$ are not available. That is why the package uses
a **high-budget reference estimate**:

$$
\tilde{\mu}_i^{(M)} = \frac{1}{M} \sum_{m=1}^M \tilde{Y}_{i,m},
$$

with a much larger budget $M$ than the finite-budget study budget $N$.

This reference is a practical stand-in for truth. It is useful because it gives
us a target for comparison:

- which action does the finite-budget method recommend?
- how much regret does that recommendation have relative to the reference-best?
- how large is the whole-vector estimation error?

But the reference is still only a proxy. If the top two actions are extremely
close, even a large-budget reference can remain uncertain. That is why the
package includes reference certification diagnostics rather than treating the
reference estimate as unquestionable fact.

# 7. Why Thompson sampling is the conceptual center

The package keeps equal allocation, greedy allocation, UCB, and OCBA as
comparators. That is important. We do not want to make Thompson sampling look
good by eliminating all alternatives.

But Thompson sampling is the conceptual center because it directly addresses the
thing this project cares about: **how uncertainty should shape rollout
allocation**.

At each step, Thompson sampling does not merely ask which action currently has
the highest posterior mean. It asks which actions are still plausibly best under
the posterior. That makes it a natural fit for finite-budget move evaluation
because the most important decisions are usually not about clearly bad moves;
they are about the plausible leaders.

# 8. What we should expect to see empirically

The project becomes more interpretable once we state, in advance, what kinds of
patterns are reasonable to expect.

## 8.1 Easy positions

When one action is clearly stronger than the rest, most sensible methods should
eventually agree. Thompson sampling may still help by discarding weak actions
faster, but the gains may be modest because the problem is easy.

## 8.2 Hard near-tie positions

These are the positions where the package becomes scientifically interesting.

When the top actions are close:

- recommendations may flip across seeds,
- finite-budget confidence can be misleading,
- adaptive allocation matters more,
- reference certification matters more.

This is where trace plots, budget studies, and finite-vs-reference comparisons
become far more informative than a single printed recommendation.

## 8.3 Large action sets with only a few plausible contenders

This is often where adaptive methods have their clearest advantage. If many
actions are visibly weak, an adaptive method can focus on the small set that
still matters.

That is not guaranteed to happen well every time, but it is the kind of
behavior the package is built to study.

# 9. What “good output” should look like

One of the practical goals of the vignette refresh is to move away from
output that feels like an internal debug dump and toward output that answers
obvious human questions:

- What decision instance am I looking at?
- How many legal moves are there?
- Which few moves should I focus on first?
- What did the method recommend?
- How much budget went into that recommendation?
- How uncertain is that recommendation?
- How did it compare with a stronger reference run?

That is why the new vignettes emphasize:

- compact action tables,
- one-row summaries,
- selected columns rather than raw internals,
- move labels rather than unexplained indices whenever possible.

# 10. Why this project generalizes beyond backgammon

The package’s results are not only about board games. The underlying lesson is
about **finite-budget adaptive simulation**.

If Thompson sampling behaves well here, that supports its use in other settings
where:

- alternatives are expensive to evaluate,
- observations are noisy,
- uncertainty should guide where the next sample goes.

Examples include:

- simulation optimization,
- A/B testing under finite traffic,
- recommender-system candidate ranking,
- hyperparameter tuning with noisy validation results,
- stochastic search over expensive design alternatives,
- sequential clinical allocation when the goal is to learn which arm is best.

The transfer is not exact because those applications differ in reward model,
cost structure, and ethical constraints. But the core idea carries over:

> when you cannot afford to sample everything equally, posterior-guided sampling
> can be a principled way to decide where to spend effort.

# 11. A tiny anchor example

The package is best read from a fixed decision instance.

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

explain_position(board, roll)
bg_print_board(board)
print(roll)

legal <- generate_legal_moves(board, roll)
summarize_legal_moves(legal, max_candidates = 6L)
```

This tiny example is doing something more important than it may first appear to
do. It turns the package from “a large set of functions” into a concrete
question:

- here is the state,
- here is the realized roll,
- here is the action set,
- now we can talk about finite-budget allocation in a precise way.

# 12. What to read next

The rest of the vignette sequence is intentionally ordered.

1. `02_thompson_sampling_theory.Rmd` explains the Thompson logic in depth.
2. `03_package_methodology.Rmd` explains how the package is engineered.
3. `04_backgammon_basics.Rmd` shows the mechanical features with readable
   outputs.
4. `05_thompson_workflow.Rmd` walks through the main Thompson-centered analysis
   slowly and connects each code block back to the statistical problem.

That sequence is meant to make the project feel less arbitrary and much more
coherent: motivation first, then theory, then engineering, then mechanics,
then the full workflow.

# 13. Experimental questions this package should answer

A useful research checklist is:

- does Thompson recover the reference-best action more often than baselines?
- how quickly does simple regret shrink as budget increases?
- does allocation concentrate on plausible leaders rather than obvious weak
  actions?
- how sensitive is behavior to hard states with tiny top-two gaps?
- how stable are recommendations over budget and over random seeds?
- how large are runtime costs relative to quality gains?

These questions are practical and measurable with the package outputs.

# 14. Strengths and limitations to state explicitly

Strengths:

- coherent finite-budget allocation framing,
- reproducible local decision instances (`board`, realized `roll`),
- explicit reference-based evaluation toolkit,
- multiple baselines for honest comparison.

Limitations:

- reference estimates are still proxy truth, not exact truth,
- conclusions are rollout-model dependent,
- near-tie states can remain ambiguous at moderate budgets,
- finite-budget instability is expected and must be reported, not hidden.

# 15. Future directions worth investigating

Likely high-value next steps:

- stronger multi-core support for benchmark grids,
- richer posterior models for bounded rollout rewards,
- expanded TTTS and stopping-rule variants,
- tighter reference-certification procedures,
- contextual/feature-informed Thompson variants for broader state families.
