---
title: "Package Methodology and Engineering Notes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Methodology and Engineering Notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# 1. Big picture: three layers, one story

The package becomes much easier to understand when it is viewed as three layers.

1. **Game mechanics layer**  
   Create boards, rolls, legal moves, and full-game simulations.

2. **Statistical allocation layer**  
   Spend a finite rollout budget across legal moves using equal allocation,
   greedy allocation, UCB, OCBA, Thompson, or TTTS.

3. **Interpretation layer**  
   Summarize results, compare with proxy truth, study budgets, and benchmark the
   methods.

The current codebase already contains these ingredients. The main challenge is
not that the algorithms are missing. The main challenge is that the user-facing
workflow has historically made the package feel more like a pile of internals
than a guided analysis pipeline.

# 2. The canonical workflow

A clean way to think about the package is:

$$
\text{board} + \text{roll}
\rightarrow \text{legal actions}
\rightarrow \text{finite-budget evaluation}
\rightarrow \text{reference comparison}
\rightarrow \text{budget study}
\rightarrow \text{many-game application}.
$$

In package terms, that is roughly:

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

legal <- generate_legal_moves(board, roll)
th <- evaluate_actions_thompson(board, roll, total_budget = 700L)

ref <- approximate_action_reference(board, roll, truth_budget = 4000L)
cmp <- compare_thompson_to_reference(
  board = board,
  roll = roll,
  total_budget = 320L,
  reference = ref
)

trade <- study_budget_tradeoff(
  board = board,
  roll = roll,
  method = "thompson",
  budgets = c(160L, 480L, 1200L),
  truth_budget = 4000L
)
```

That is the methodological spine of the project.

# 3. Why the current workflow can feel confusing

After auditing the code and existing vignettes, the biggest problems are not in
the algorithms themselves. They are in how information is surfaced.

## 3.1 Too many layers are exposed at once

The package has:

- core mechanics,
- statistical wrappers,
- research helpers,
- print methods,
- standalone docs,
- overlapping vignettes.

All of those pieces are individually reasonable, but together they can make it
hard to see which route is the *main* route for a typical user.

## 3.2 Some printed objects are too raw

The most obvious example is `print.bg_game_result()`, which historically dumped
the entire game history by default. The raw history is useful for debugging but
not ideal for explanation. A reader usually wants:

- who played whom,
- how many turns occurred,
- who won,
- a short readable head of the history.

That is why the refreshed vignettes avoid giant raw dumps and prefer compact
views.

## 3.3 Several names are conceptually similar

There are user-facing aliases in `R/statistical_api.R`, deeper research helpers
in `R/thompson_research.R`, study wrappers in `R/statistical_studies.R`, and
other convenience layers in `R/reporting.R` and `R/recommendation.R`.

Those layers are not wrong, but they can obscure the canonical path. A reader
should not need to guess whether a problem belongs to:

- `evaluate_actions_thompson()`,
- `evaluate_moves_thompson()`,
- `bg_analysis_report()`,
- `bg_recommend_move()`.

A methodology document should make those relationships explicit.

# 4. R-layer design: what each file family is doing

This section describes the package the way a maintainer or advanced user should
read it.

## 4.1 Core mechanics wrappers

Files such as `board.R`, `dice.R`, `move.R`, `legal_moves.R`, and `game.R`
handle the mechanics:

- board construction and validation,
- dice representation,
- move representation,
- legal-move generation,
- full-turn and full-game progression.

This is the layer that turns backgammon from an abstract topic into a concrete
decision environment.

## 4.2 Statistical API layer

`R/statistical_api.R` is trying to provide more research-friendly names and a
shallower learning curve:

- `initialize_board()`,
- `roll_dice()`,
- `generate_legal_moves()`,
- `simulate_game()`,
- `evaluate_moves_thompson()`,
- `compare_action_posteriors()`.

This is a helpful layer for pedagogy, but it also creates alias duplication.
That is manageable as long as the documentation clearly states which names are
canonical and which are convenience wrappers.

## 4.3 Allocation methods

`R/allocation_methods.R` is the central statistical layer for one-position move
evaluation. It provides:

- equal allocation,
- greedy allocation,
- UCB,
- OCBA,
- Thompson,
- TTTS,
- proxy-truth approximation,
- multi-case benchmarking support.

This file is the natural center of the move-evaluation story.

## 4.4 Thompson-specific research helpers

`R/thompson_research.R` adds higher-level objects that are very useful for the
project’s stated goals:

- allocation tracing,
- finite-vs-reference comparison,
- reference certification,
- Thompson-focused benchmarking.

These functions make the package feel like a research workflow rather than only
a move-evaluation engine.

## 4.5 Study wrappers and reporting

`R/statistical_studies.R` and `R/reporting.R` sit above the lower-level
evaluation functions and help turn repeated patterns into readable objects.

These are helpful, but they are also part of why the project benefits from a
clear methodology document. Once there are many convenience layers, readers need
to know which ones are intended for explanation, which ones are intended for
experimentation, and which ones are mostly internal glue.

# 5. C++ layer comments: what is strong and what is worth revisiting

The C++ layer is generally the right place for the heavy loops. The audit did
not suggest that the engine is conceptually broken. In fact, several pieces are
quite well structured. The main opportunities are around clarity, consolidation,
and future optimization.

## 5.1 `bg_board.*`

This module is the board representation backbone. It is doing the right kind of
work: validate, clone, and normalize board data close to the engine.

**Comment:** this file looks like a good foundation and should remain boring.
That is a compliment. Stable representation code is exactly what the rest of the
system needs.

## 5.2 `bg_rules.*`

This is the legality layer for checker movement and rule enforcement.

**Comment:** keeping the rule primitives separate from higher-level statistical
code is a strong design choice. It reduces the chance that move-generation logic
and rollout-allocation logic become tangled.

## 5.3 `bg_move.*`

This module handles move-step and move-sequence representations.

**Comment:** this separation is useful because it lets the package expose move
labels and move summaries cleanly at the R layer. It is a bridge between
mechanics and interpretation.

## 5.4 `bg_movegen.*`

This module is one of the most important mechanics files because it creates the
candidate action set.

**Comment:** the move-generation and random-legal-sequence logic here appears
performance-minded in a good way. It deserves to remain separate. For this
project, efficient candidate construction matters because every higher-level
study begins here.

## 5.5 `bg_game.*`

This file orchestrates turn and game simulation and dispatches between player
types.

**Comment:** it is doing important integration work, but policy dispatch logic
can become crowded over time. If the package grows, this file is one of the
first places where complexity could accumulate.

## 5.6 `bg_rollout.*`

This module is comparatively thin and mostly handles rollout configuration.

**Comment:** it makes sense as a boundary file, but it is also the kind of place
where duplication can hide. Thin glue is fine, but only if it is actually
reducing complexity.

## 5.7 `bg_allocation.*`

This is the statistical heart of the engine.

It handles:

- adaptive rollout allocation,
- posterior bookkeeping,
- diagnostics,
- trace support,
- candidate collapsing,
- variance-control options.

**Comment:** centralizing the hard allocation logic here is the right idea, but
this is also the file most likely to surprise users indirectly. In particular,
the candidate-collapsing logic is mathematically reasonable and performance
sensible, yet it is the source of one of the most confusing user-facing
artifacts: displayed candidate indices can refer to representative original
moves rather than contiguous printed row numbers.

That is not a reason to rewrite the logic immediately, but it is exactly the
kind of implementation detail that needs documentation or a cleaner display
layer.

## 5.8 `bg_simulation.*`

This file powers repeated game simulation and compact matchup summaries.

**Comment:** this is a strength of the codebase. Repeated-game experiments should
not be implemented by calling full game simulation from R in a loop. The C++
path here is the right engineering decision and makes the many-game vignettes
possible.

## 5.9 `bg_benchmark.*`

This module exists to support repeated case-level benchmarking.

**Comment:** conceptually useful, but benchmark grids are also an obvious future
parallelization target because many cells are independent.

## 5.10 `bg_thompson_rollout.*`

This is a thin Thompson-oriented adapter layer.

**Comment:** this file feels like a candidate for consolidation or at least a
documentation note explaining why it exists separately. Thin wrappers are fine
if they make the public interface clearer; they are less useful if they merely
duplicate a subset of `bg_allocation.*`.

## 5.11 `bg_random_player.*` and `bg_heuristic.*`

These modules provide baseline player behavior.

**Comment:** they are simple but essential. They make it possible to compare
Thompson-driven play against easier baselines and to define rollout policies
inside continuation playouts.

# 6. Engineering choices that matter for interpretation

Several implementation decisions are not just engineering details. They affect
how results should be read.

## 6.1 Candidate collapsing

Equivalent legal move sequences that lead to the same post-move board can be
collapsed inside the allocation engine.

That is computationally sensible. But it introduces a display problem:

- the printed action table may have fewer rows than the original raw move list,
- representative indices can look “skipped” or unexpectedly large.

This is one of the most important sources of confusion for users and should be
handled more gracefully in the display layer.

## 6.2 Fast diagnostics versus full diagnostics

Some objects support `fast_diagnostics = TRUE`, which skips probability-best and
expected-regret calculations.

That is a good runtime tradeoff, but it means the same printed object can look
different depending on settings. Documentation should say that explicitly.

## 6.3 Proxy truth is itself computed, not given

Reference truth is approximate and simulated. The package handles this honestly
through certification helpers, but readers should be reminded that:

- the reference is stronger than the finite-budget run,
- the reference is not exact,
- small top-two reference gaps should be interpreted carefully.

# 7. Room for improvement without changing the science

The package is already doing interesting work. The most useful next steps are
not radical algorithm changes. They are clarity and engineering improvements.

## 7.1 Cleaner user-facing outputs

A good next pass would standardize a small set of compact display tables across
the package so that every major object answers the same few questions:

- what was recommended?
- how much budget was spent?
- how uncertain is the recommendation?
- what should I compare it against?

The existing compact-table helpers are already moving in this direction.

## 7.2 Clearer canonical names

It would help to define one canonical name for each task and treat other names
as convenience aliases in the documentation, rather than letting all names
appear equally primary.

## 7.3 Reduce documentation overlap

Before the vignette refresh, the package had multiple documents covering similar
ground with slightly different framing. That made it harder to identify the
source of truth.

The new vignette sequence is intended to fix that by making the vignettes the
main narrative and the older standalone notes clearly secondary.

# 8. Multi-core opportunities

The user specifically asked about room for improvement through more cores. This
is one of the clearest future optimization paths.

## 8.1 Benchmark-grid parallelism

Many benchmark cells are independent:

- different cases,
- different budgets,
- different methods,
- different variance-control settings.

That kind of workload is an obvious candidate for multi-core execution at the R
layer using packages like `parallel`, `future.apply`, or a similar abstraction.

## 8.2 Repeated-game simulation parallelism

Large repeated matchups are also naturally parallel at the replication level.
Each game can often be simulated independently once the seeding strategy is
carefully defined.

## 8.3 Allocation-level parallelism

Within one move-evaluation problem, there may also be room to parallelize some
rollout work across candidate actions or across posterior-diagnostic draws.
This is more delicate because Thompson-style allocation is sequential, but there
are still pieces of the surrounding workload that could be parallelized.

## 8.4 Caveats

Parallelism is not free. This package uses seeds, optional CRN, and dice-control
settings. Any multi-core implementation would need to preserve:

- reproducibility,
- sensible seed derivation,
- CRN semantics where requested,
- avoidance of thread oversubscription.

That is why the methodology vignette discusses multi-core as a real opportunity
without treating it as a trivial patch.

# 9. A small methodology example

The code below shows the intended progression without trying to do everything at
once.

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

# Mechanics first.
legal <- generate_legal_moves(board, roll)
summarize_legal_moves(legal, max_candidates = 6L)

# Then one finite-budget evaluation.
th <- evaluate_actions_thompson(
  board = board,
  roll = roll,
  total_budget = 700L,
  rollout_policy = "random",
  max_rollout_turns = 250L,
  fast_diagnostics = FALSE,
  seed = 1L
)

summary(th)

# Then reference comparison.
ref <- approximate_action_reference(
  board = board,
  roll = roll,
  truth_budget = 4000L,
  rollout_policy = "random",
  max_rollout_turns = 250L,
  seed = 1L
)

cmp <- compare_thompson_to_reference(
  board = board,
  roll = roll,
  total_budget = 320L,
  reference = ref,
  rollout_policy = "random",
  max_rollout_turns = 250L,
  seed = 2L
)

cmp$summary
```

That sequence is methodological, not just mechanical. It tells the user what the
package is for.

# 10. Final engineering takeaway

The codebase already contains the important ingredients:

- a solid mechanics layer,
- a serious allocation engine,
- useful research helpers,
- fast repeated simulation,
- compact table helpers.

What it needed most was a clearer narrative and a cleaner route through the
package. That is why the new documentation prioritizes coherence over sheer
function coverage.
