---
title: "Backgammon Basics with Simple Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backgammon Basics with Simple Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette is intentionally modest. The goal is not to show every feature.
The goal is to make the basic mechanics readable.

# 1. Set up one decision instance

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

explain_position(board, roll)
```

That sentence is a good way to anchor everything else. We are studying one board
state and one realized roll.

# 2. Print the board and the roll

```{r}
bg_print_board(board)
print(roll)
```

Read this output first, before looking at any allocation method. If the board
and roll are not clear, the rest of the analysis will feel abstract.

# 3. List a few legal moves

```{r}
legal <- generate_legal_moves(board, roll)

legal_summary <- summarize_legal_moves(
  legal,
  max_candidates = 6L,
  print_table = FALSE
)

legal_summary
```

Why this output is friendly:

- it shows only a few columns,
- the move label is easier to read than a deeply nested internal object,
- `max_candidates` keeps the display manageable.

# 4. Apply one move and inspect the result

```{r}
move1 <- legal[[1L]]
board_after <- apply_move(board, move1)

cat("MOVE\n")
print(move1)

cat("\nBEFORE\n")
bg_print_board(board)

cat("\nAFTER\n")
bg_print_board(board_after)
```

This is one of the most useful beginner checks in the package. It turns a move
label into a concrete state transition.

# 5. Simulate one short game without printing everything

A single game is useful for intuition, but the full raw history can be more than
you need at first. So the safest pattern is:

1. simulate the game,
2. print a one-row overview,
3. inspect only the first few history rows.

```{r}
one_game <- simulate_game(
  state = board,
  policy_white = "random",
  policy_black = "random",
  max_turns = 80L,
  seed = 1L
)

game_overview <- data.frame(
  player_1 = one_game$player1_selection,
  player_2 = one_game$player2_selection,
  n_turns = one_game$n_turns,
  game_over = one_game$game_over,
  winner = one_game$winner,
  turn_limit_reached = one_game$turn_limit_reached,
  stringsAsFactors = FALSE
)

game_overview

head(
  one_game$history[, c(
    "turn", "player", "selection", "die1", "die2",
    "n_legal_moves", "chosen_n_steps", "game_over", "winner"
  )],
  8L
)
```

This keeps the output readable while still showing how turn-level simulation
works.

# 6. Run many games and summarize them compactly

When the goal is repeated simulation, `simulate_matchup()` is usually a better
tool than calling `simulate_game()` in a loop. It returns one row per game plus
a compact summary.

```{r}
random_matchup <- simulate_matchup(
  player1 = "random",
  player2 = "random",
  n_games = 100L,
  board = board,
  max_turns = 150L,
  seed = 2L
)

summary(random_matchup)[, c(
  "n_games",
  "completed_games",
  "player1_wins",
  "player2_wins",
  "player1_win_rate",
  "player2_win_rate",
  "mean_turns"
)]

head(
  random_matchup$games[, c(
    "game_id",
    "winner_label",
    "n_turns",
    "game_over",
    "turn_limit_reached"
  )],
  8L
)
```

This is a much better first look than a giant raw table.

# 7. Use a stronger player only after the basics are clear

Once the mechanics are comfortable, you can compare a rollout-based player
against a simple baseline.

```{r}
thompson_matchup <- simulate_matchup(
  player1 = "thompson_rollout",
  player2 = "random",
  n_games = 100L,
  board = board,
  max_turns = 150L,
  rollout_budget = 16L,
  rollout_policy = "random",
  max_rollout_turns = 150L,
  seed = 3L
)

summary(thompson_matchup)[, c(
  "n_games",
  "completed_games",
  "player1_wins",
  "player2_wins",
  "player1_win_rate",
  "player2_win_rate",
  "mean_turns"
)]
```

This is a good bridge from mechanics to the later statistical vignettes:
a local allocation policy becomes a player only when it is used repeatedly in
full games.

# 8. What this vignette deliberately avoided

This vignette did **not** try to show:

- every column available in every object,
- giant history tables,
- all move-evaluation methods,
- all benchmark helpers.

That is intentional. The point here is to make the package feel easy to read at
the mechanics level. The heavier statistical workflow belongs in the next
vignette.
