---
title: "Easy Function Calls: Setup + Step-by-Step Workflow"
output: rmarkdown::html_vignette
params:
  run_code: false
  quick_mode: true
vignette: >
  %\VignetteIndexEntry{Easy Function Calls: Setup + Step-by-Step Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
if (!exists("params")) {
  params <- list(run_code = FALSE, quick_mode = TRUE)
}
run_code <- isTRUE(params$run_code)
options(bg_quick_mode = isTRUE(params$quick_mode))

fmt_elapsed <- function(tm) round(unname(tm[["elapsed"]]), 3)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_code,
  fig.width = 7.0,
  fig.height = 4.5
)
```

This document is a slow, practical walkthrough of the main user-facing calls.

- It starts with package initialization from GitHub.
- It then runs one function at a time.
- Each section explains what question the function answers, which parameters matter,
  and what output to inspect.
- Timing (`system.time()`) is included so you can judge runtime costs.

Expected runtime if you render with `run_code = TRUE`:

- quick mode (`quick_mode = TRUE`): roughly 1 to 4 minutes on a laptop.
- deep mode (`quick_mode = FALSE`): roughly 6 to 20+ minutes.

# 0. Install and initialize the package

If you are starting from GitHub, use this first. This chunk is shown with
`eval = FALSE` so it does not reinstall every render.

```{r, eval = FALSE}
# Install helper once if needed.
install.packages("remotes")

# Install backgammonr from GitHub.
remotes::install_github("gabrielalwan/backgammonr")

# Load package and verify.
library(backgammonr)
packageVersion("backgammonr")
```

If you are inside the package source folder in RStudio and want local
development mode:

```{r, eval = FALSE}
install.packages("devtools")
devtools::load_all(".")
```

# 1. `bg_initial_board()`

`bg_initial_board()` builds a standard starting board object. Use it when you
want a reproducible, known state before testing any simulation method.

```{r}
library(backgammonr)

tm <- system.time({
  board <- bg_initial_board()
})

bg_print_board(board)

data.frame(
  function_call = "bg_initial_board",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 2. `bg_roll()`

`bg_roll(d1, d2)` creates a realized dice roll object. In examples, fixed dice
make results reproducible and easier to interpret than random draws.

```{r}
library(backgammonr)

tm <- system.time({
  roll <- bg_roll(1L, 6L)
})

print(roll)

data.frame(
  function_call = "bg_roll",
  die_1 = 1L,
  die_2 = 6L,
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 3. `generate_legal_moves()`

`generate_legal_moves(board, roll)` defines the action set for this decision.
This is the candidate list over which Thompson (or any baseline) allocates
finite simulation budget.

```{r}
library(backgammonr)

board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

tm <- system.time({
  legal <- generate_legal_moves(board, roll)
  legal_view <- summarize_legal_moves(legal, max_candidates = 10L, print_table = FALSE)
})

legal_view[, c("candidate_index", "move_label", "n_steps")]

data.frame(
  function_call = "generate_legal_moves",
  n_candidates = nrow(legal_view),
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 4. `evaluate_actions_thompson()`

`evaluate_actions_thompson()` is the central finite-budget call. It allocates
rollouts adaptively using posterior sampling and returns recommendation,
allocation, estimated value, uncertainty, and probability-best style summaries.

Key parameters to tune:

- `total_budget`: total rollout budget available now.
- `rollout_policy`: policy used during rollout simulation.
- `max_rollout_turns`: rollout truncation limit.
- `seed`: reproducibility for stochastic simulation.

```{r}
library(backgammonr)

quick <- isTRUE(getOption("bg_quick_mode", TRUE))
total_budget <- if (quick) 800L else 2000L
max_rollout_turns <- if (quick) 220L else 300L
rollout_policy <- "random"
seed <- 11L

params_eval <- data.frame(
  total_budget = total_budget,
  rollout_policy = rollout_policy,
  max_rollout_turns = max_rollout_turns,
  seed = seed
)
params_eval

tm <- system.time({
  th <- evaluate_actions_thompson(
    board = bg_initial_board(),
    roll = bg_roll(1L, 6L),
    total_budget = total_budget,
    rollout_policy = rollout_policy,
    max_rollout_turns = max_rollout_turns,
    fast_diagnostics = FALSE,
    seed = seed
  )
})

summary(th)[, c(
  "method", "total_budget", "recommended_move_label",
  "recommended_estimate", "recommended_prob_best",
  "recommended_expected_regret", "recommended_allocation_count",
  "runtime_seconds"
)]

th[["results"]][, c(
  "move_label", "recommended", "allocation_count",
  "estimate", "posterior_sd", "prob_best", "posterior_expected_regret"
)]

data.frame(
  function_call = "evaluate_actions_thompson",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 5. `approximate_action_reference()`

`approximate_action_reference()` builds a stronger high-budget reference estimate
for the same board and roll. This is proxy truth, not exact truth, and should
be interpreted with the reference certificate.

```{r}
library(backgammonr)

quick <- isTRUE(getOption("bg_quick_mode", TRUE))
truth_budget <- if (quick) 5000L else 20000L
max_rollout_turns <- if (quick) 220L else 300L
seed <- 21L

params_ref <- data.frame(
  truth_budget = truth_budget,
  rollout_policy = "random",
  max_rollout_turns = max_rollout_turns,
  seed = seed
)
params_ref

tm <- system.time({
  ref <- approximate_action_reference(
    board = bg_initial_board(),
    roll = bg_roll(1L, 6L),
    truth_budget = truth_budget,
    rollout_policy = "random",
    max_rollout_turns = max_rollout_turns,
    seed = seed
  )

  cert <- certify_reference_truth(reference = ref)
})

ref[["results"]][1:8, c(
  "move_label", "allocation_count", "estimate", "posterior_sd", "recommended"
)]

cert[["certificate"]][, c(
  "reference_best_move_label",
  "reference_second_move_label",
  "top_two_gap_estimate",
  "top_two_gap_lower_95",
  "top_two_gap_upper_95",
  "certified",
  "difficulty_label"
)]

data.frame(
  function_call = "approximate_action_reference",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 6. `compare_thompson_to_reference()`

This function answers the core practical question: with a finite budget, how
close is Thompson to the high-budget reference?

Important detail: pass `reference` and `reference_certificate` if you already
computed them. That avoids re-running expensive reference simulations.

```{r}
library(backgammonr)

quick <- isTRUE(getOption("bg_quick_mode", TRUE))
board <- bg_initial_board()
roll <- bg_roll(1L, 6L)

# Build once, then reuse.
ref <- approximate_action_reference(
  board = board,
  roll = roll,
  truth_budget = if (quick) 5000L else 20000L,
  rollout_policy = "random",
  max_rollout_turns = if (quick) 220L else 300L,
  seed = 31L
)
cert <- certify_reference_truth(reference = ref)

params_cmp <- data.frame(
  method = "thompson",
  total_budget = if (quick) 1000L else 2500L,
  reference_budget = ref$truth_budget,
  rollout_policy = "random",
  seed = 32L
)
params_cmp

tm <- system.time({
  cmp <- compare_thompson_to_reference(
    board = board,
    roll = roll,
    method = "thompson",
    total_budget = if (quick) 1000L else 2500L,
    reference = ref,
    reference_certificate = cert,
    rollout_policy = "random",
    max_rollout_turns = if (quick) 220L else 300L,
    seed = 32L
  )
})

cmp[["summary"]][, c(
  "method", "total_budget", "reference_budget",
  "chosen_move_label", "reference_best_move_label",
  "proxy_pcs", "simple_regret", "mse",
  "reference_certified", "difficulty_label",
  "finite_runtime_seconds", "reference_runtime_seconds"
)]

cmp[["action_table"]][, c(
  "move_label", "finite_recommended", "reference_best",
  "finite_allocation_count", "finite_estimate", "finite_prob_best",
  "reference_estimate", "estimation_error", "abs_error"
)][1:8, ]

data.frame(
  function_call = "compare_thompson_to_reference",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 7. `study_budget_tradeoff()`

`study_budget_tradeoff()` repeats the finite-budget study over a budget grid.
Use it to see how regret/MSE/selection improve as you add simulation budget.

```{r}
library(backgammonr)

quick <- isTRUE(getOption("bg_quick_mode", TRUE))
budgets <- if (quick) c(128L, 256L, 512L, 1024L) else c(128L, 256L, 512L, 1024L, 2048L)

params_trade <- data.frame(
  budgets = paste(budgets, collapse = ", "),
  truth_budget = if (quick) 5000L else 20000L,
  rollout_policy = "random",
  seed = 41L
)
params_trade

tm <- system.time({
  trade <- study_budget_tradeoff(
    board = bg_initial_board(),
    roll = bg_roll(1L, 6L),
    method = "thompson",
    budgets = budgets,
    truth_budget = if (quick) 5000L else 20000L,
    rollout_policy = "random",
    max_rollout_turns = if (quick) 220L else 300L,
    seed = 41L
  )
})

trade[["results"]][, c(
  "total_budget", "chosen_move_label", "truth_best_move_label",
  "correct_selection", "simple_regret", "mse", "runtime_seconds"
)]

data.frame(
  function_call = "study_budget_tradeoff",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 8. `study_variance_controls()`

`study_variance_controls()` studies finite-budget sensitivity to simulation
design controls (dice stratification and CRN). Use it to test robustness.

```{r}
library(backgammonr)

quick <- isTRUE(getOption("bg_quick_mode", TRUE))

params_var <- data.frame(
  total_budget = if (quick) 900L else 1800L,
  dice_modes = "iid, stratified_first_roll",
  crn_values = "FALSE, TRUE",
  truth_budget = if (quick) 5000L else 20000L,
  seed = 51L
)
params_var

tm <- system.time({
  var_study <- study_variance_controls(
    board = bg_initial_board(),
    roll = bg_roll(1L, 6L),
    method = "thompson",
    total_budget = if (quick) 900L else 1800L,
    dice_modes = c("iid", "stratified_first_roll"),
    crn_values = c(FALSE, TRUE),
    truth_budget = if (quick) 5000L else 20000L,
    rollout_policy = "random",
    max_rollout_turns = if (quick) 220L else 300L,
    seed = 51L
  )
})

var_study[["results"]][, c(
  "dice_mode", "crn", "chosen_move_label",
  "truth_best_move_label", "correct_selection",
  "simple_regret", "mse", "runtime_seconds"
)]

data.frame(
  function_call = "study_variance_controls",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 9. `summarize_thompson_benchmark()`

This benchmark summary is the multi-case view. It asks whether Thompson keeps
an advantage over baselines across cases and budgets, not just in one position.

```{r}
library(backgammonr)

quick <- isTRUE(getOption("bg_quick_mode", TRUE))

cases <- list(
  bg_benchmark_case(bg_initial_board(), bg_roll(1L, 6L), case_id = "init_1_6"),
  bg_benchmark_case(bg_initial_board(), bg_roll(3L, 2L), case_id = "init_3_2"),
  bg_benchmark_case(bg_initial_board(), bg_roll(5L, 4L), case_id = "init_5_4")
)

params_bm <- data.frame(
  n_cases = length(cases),
  budgets = paste(if (quick) c(256L, 1024L) else c(256L, 1024L, 4096L), collapse = ", "),
  baselines = "equal, ucb, ocba, greedy",
  reference_budget = if (quick) 5000L else 20000L,
  seed = 61L
)
params_bm

tm <- system.time({
  bm <- benchmark_thompson(
    cases = cases,
    budgets = if (quick) c(256L, 1024L) else c(256L, 1024L, 4096L),
    baselines = c("equal", "ucb", "ocba", "greedy"),
    include_ttts = TRUE,
    reference_budget = if (quick) 5000L else 20000L,
    rollout_policy = "random",
    max_rollout_turns = if (quick) 220L else 300L,
    seed = 61L
  )

  focus <- summarize_thompson_benchmark(bm)
})

focus[["thompson"]][, c(
  "total_budget", "probability_correct_selection",
  "mean_simple_regret", "mean_mse", "mean_runtime_seconds",
  "reference_best_move_label", "most_selected_move_label"
)]

head(focus[["relative_to_thompson"]], 8L)[, c(
  "baseline_method", "total_budget",
  "thompson_advantage_pcs",
  "thompson_advantage_regret",
  "thompson_advantage_mse",
  "thompson_runtime_ratio"
)]

data.frame(
  function_call = "summarize_thompson_benchmark",
  elapsed_seconds = fmt_elapsed(tm)
)
```

# 10. Recommended run order

A practical learning/testing sequence:

1. Initialize package and verify version.
2. Create board and roll (`bg_initial_board()`, `bg_roll()`).
3. Enumerate actions (`generate_legal_moves()`).
4. Run finite Thompson (`evaluate_actions_thompson()`).
5. Build one reusable high-budget reference (`approximate_action_reference()`).
6. Compare finite vs reference (`compare_thompson_to_reference()`).
7. Add budget sensitivity (`study_budget_tradeoff()`).
8. Add variance-control sensitivity (`study_variance_controls()`).
9. Finish with benchmark summary (`summarize_thompson_benchmark()`).
